\speculake: Hiding Malware in Speculative Execution

### 2018-08-02 paper/discussion.tex

\subsection{Defenses}
We now address possible defenses to detecting and reverse engineering malware that uses
\speculake.

@@

Finally, \speculake could also use another side channel method that
avoids the cache
to exfiltrate information from the speculative gadgets, such as the branch
predictor itself~\cite{evtyushkin2018branchscope}, memory bandwidth, or power
utilization of the system.

@@

\subsection{Resilience of \speculake}
%Spectre works to use a cache side-channel to leak information from a process. As
%such defenses to Spectre work to limit the information available to a
%speculatively executing program. 
%

\subsubsection{Gadget Location}
The benefits of \speculake rely on the computation performed in the
speculative function being well hidden. To this end we have verified 
certain requirements about the locations that a speculative gadget can 
be placed within a binary.

\subsubsection{Hypervisors}
For virtual environments that provide guests significant periods of access to 
hardware resources, we have found that \speculake is viable in certain configurations.

\texttt{Guest1 -> Guest1}---When both the trigger and payload are run internal to a
single virtual machine the branch prediction is successfully trained in the
payload program, though at a notably reduced hit rate. This can be overcome by
using an increased number of iterations to establish confidence.

\texttt{Guest -> Host}---In a configuration where the trigger runs on the guest
while the payload runs on the host we have found that \speculake also
successfully trains the branch predictor well enough retrieve a signal. Again,
this signal is significantly reduced, implying the hypercall context switch from
guest to host requires some overhead. We note that processor affinity in a guest
is not mapped to processor affinity in the host, as the host assigns the VM to
any available CPU, though eventually the hypervisor will assign the VM back to
the same core as the payload program allowing progress to continue.

\texttt{Host -> Guest}---When the trigger program executes on the host
attempting to train the
branch predictor to influence a payload process in a guest we have found that
the payload receives no signal. This implies that the hypercall context switch
from the host to the guest is heavier and has a larger impact on branch history.
This is also supported by the fact that hypervisors can be fingerprinted from 
within a VM using the branch predictor as demonstrated by the Project Zero 
team at Google~\cite{}.
~\cite{project_zero}.

\texttt{Guest1 -> Guest2}---Configurations where the trigger and payload run on
co-resident virtual machines similarly show no signal for the same reasons that
host to guest \speculake fails. However, this has the added difficulty that
virtual machines must be run on the same core or on a pair of hyperthreaded
cores.

>>>>

\subsection{Resilience of \speculake}
%Spectre works to use a cache side-channel to leak information from a process. As
%such defenses to Spectre work to limit the information available to a
%speculatively executing program. 
%
\subsubsection{Spectre Versions}
In this work we have implemented applications making use of the speculative
primitive in Spectre version 2, however the speculative primitive is not 
unique to this varaint. The \speculake model can be applied more broadly
across the spectrum of speculative functionality to hide malicious computation.

For example an attacker could construct a series of direct branches with 
contradictory conditions such that the resulting code section can never 
be on a real world execution path.
This, however does not prevent the section from being speculated. An 
attacker may be able to hide a jump to a speculative entry point in 
this section and train the branch predictor to take this path.

Alternatively, recent variants of Spectre version 1 could be used to
redirect control flow in speculative overflow style attacks~\cite{}. 
Using this technique speculative gadgets could be executed
based on attacker controlled input.

Similarly we theorize that speculative-store-bypass could be used to
redirect control flow when the mis-speculated load is that of an indirect
branch target or function pointer. Again the speculative primitive could 
be employed to execute speculative gadgets.

These alternatives create an even larger attack area for a potential 
reverse engineer as the attacker can pivot between them with minimal alterations.

\subsubsection{Gadget Location}
The benefits of \speculake rely on the computation performed in the
speculative world being well hidden. To this end we have verified 
certain requirements about the locations that a speculative gadget can 
be placed within a binary.

\subsubsection{Hypervisors}
For virtual environments that provide guests significant periods of access to
hardware resources, \speculake is viable in certain configurations. 
This result is specific to Spectre version 2 as it relies on 
the indirect branch predictor effecting the branches in either the host 
or guest environment. We have found that the hypercall context switch from 
guest to host is light enough that a trigger program running in a guest
can effect the branch predictor in a host on the same CPU core. This is not
true in the alternate configuration where the trigger runs on the host as the 
hypercall transition seems to be more substantial. This implies, and we have verified,
that guest-to-guest indirect branch predictor interference, and the current
\speculake implementation, will not work across disparate virtual environments.

\subsection{Defenses}
We now address possible defenses to detecting and reverse engineering malware that uses
\speculake. 
\subsubsection{Implemented Mitigations}
Multiple mitigation have been implemented in production and consumer environments
already, stretching from compiler patches to micro-code updates.

IBPB - IBPB is used when transitioning to a new secure address space allowing a
program to ensure that earlier code's behavior does not effect its branch prediction.
However, this does not prevent a willing payload program from intentionally omitting
IBPB protections.

STIBP - As discussed in section~\ref{sssec:hyperthread} Single Thread Indirect Branch Predictors
prevent sibling threads from interacting via the Indirect Branch Predictor~\cite{}.
However, this does not prevent coresident processes from cooperating, which is
guaranteed to eventually occur in most modern operating systems.

IBRS - Prevents code from less privileged prediction modes from effecting predicted
addresses, specifically used for portecting the kernel. This does not
prevent speculative execution in a willing payload program in a less privileged
speculation mode.

Retpoline - Using speculation to prevent speculation, retpoline is a construction
used to prevent arbitrary indirect jumps that can be included inline by a compiler
or patched into a binary~\cite{}. While this successfully thwarts indirect branch prediction
tampering it incurs a siginificant performance penalty when applied globally to all
indirect branches~\cite{}.

@@

Finally, \speculake could also use another side channel method that
avoids the cache
to exfiltrate information from the speculative gadgets, such as the branch
predictor itself~\cite{evtyushkin2018branchscope}, memory bandwidth, power
utilization, or contention over other shared resources. While cache 
channels tend to have the highest throughput, they are not the only 
resource that must be monitored to detect or prevent these types of attacks.

@@

### 2018-08-02 paper/introduction.tex

Previously, it was assumed that speculative execution results remain invisible
if discarded, as careful CPU design maintains strict separation between
speculative results and updates to architectual state. However, recent research
has revealed side channels that violate this separation, and researchers have
demonstrated ways to exfiltrate results from speculative computation. Most
notably, the Spectre vulnerability allows attackers to leak information from
purposefully mis-speculated branches in a victim process~\cite{spectre}. The
Meltdown vulnerability uses speculative results of an unauthorized memory read
to sidestep page faults and leak protected memory from the
kernel~\cite{meltdown}. Both of these vulnerabilities focus on extracting secret
data from a process or operating system. Recent follow-up work has revealed
other Spectre ``variants'',  including speculative buffer overflows, speculative
store bypass, and using alternative side channels besides the
cache~\cite{kiriansky2018speculative,spectre3}. In addition, several attacks
have leveraged Spectre to attack Intel's SGX~\cite{chen2018sgxpectre,
spectre_sgx, foreshadow}, and perform remote leakage
attacks~\cite{schwarz2018netspectre}.

>>>>

Previously, it was assumed that speculative execution results remain invisible
if discarded, as careful CPU design maintains strict separation between
speculative results and updates to architectual state. However, recent research
has revealed side channels that violate this separation, and researchers have
demonstrated ways to exfiltrate results from speculative computation. Most
notably, the Spectre vulnerability allows attackers to leak information from
purposefully mis-speculated branches in a victim process~\cite{spectre}. The
Meltdown vulnerability uses speculative results of an unauthorized memory read
to sidestep page faults and leak protected memory from the
kernel~\cite{meltdown}. Both of these vulnerabilities focus on extracting secret
data from a process or operating system. Recent follow-up work has revealed
other Spectre ``variants'',  including speculative buffer overflows, speculative
store bypass, and using alternative side channels besides the
cache~\cite{kiriansky2018speculative, spectre_ng}. In addition, several attacks
have leveraged Spectre to attack Intel's SGX~\cite{chen2018sgxpectre,
spectre_sgx, foreshadow}, and perform remote leakage
attacks~\cite{schwarz2018netspectre}.
